A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-05-23 manuscript/manuscript.Rmd
Coming along

output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

@@

The ability to accurately forecast water temperature, dissolved oxygen and chlorophyll concentration in freshwater systems is important to natural resource management. Since water temperature, dissolved oxygen and chlorophyll concentration effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive forecasts of these variables could allow managers to prepare for scenarios with far-reaching negative impacts. Limnologists have historically used empirical and process-based models to forecast water quality metrics, but in recent years, researchers have shown that machine learning can outperform these traditionally used methods \cite{Hanson, others}. A shortcoming of the limnological studies that support machine learning is that they tend to focus on a handful of sites that are not necessarily representative of water systems on a broad geographic scale. This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by examining how these models perform at 34 different sites across North America at different times of the year.

The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}. A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}. For instance, in a model comparison that examined the forecasts for phenology time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model. Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the climatology model.

There are many neural network architectures that have not been explored for ecological time series forecasting. Past applications of machine learning to limnological time series forecasting have typically used the long-short-term-memory (LSTM) neural network which was first introduced in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, there has been much development in the subfield of machine learning for time series forecasting, but recent studies examining machine learning for ecological time series forecasting have not explored more contemporary neural network architectures. This paper will compare 8 machine learning algorithms including LSTM's to ML methods that have been recently developed for time series forecasting like Temporal Fusion Transformer's and Temporal Convolutional Networks \cite{tft, tcn}.

@@

We employed ML models to generate water temperature, dissolved oxygen and chlorophyll-A concentration forecasts for 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON. 8 different ML models were trained for every site and target variable available in NEON Ecological Forecasting Challenge Aquatics theme. For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates. For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate. The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2018 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

We compared the ML models to a climatology and naive persistence null model. The climatology model generates forecasts by finding the daily mean and standard deviation of the target time series, and then drawing samples from a Gaussian Distribution with these parameters. The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast window.

The imputation of missing data proved to be critical to the performance of the machine learning models. After experimenting with data filling methods that resulted in poor model performance, we settled on a imputation method inspired by the climatology model. Missing data is estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly, seasonal or global median was used with preference given to the median available from the shortest timescale.

@@

>>>>

output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---
\setlength{\parindent}{15pt}

@@

The ability to accurately forecast water quality variables has become increasingly important in the era of global change. Freshwater ecosystems have been disproportionately impacted by anthropogenic activities, a trend that is expected to continue in the future \cite{albert2021}. Since water quality variables, like water temperature, dissolved oxygen and chlorophyll concentration, effect many biological and physical processes in freshwater ecosystems \cite{oullet-proulx, Stajkowski, Chen}, preemptive water quality forecasts could allow managers to evade situations with far-reaching negative consequences \cite{}. Limnologists have historically used statistical, process-based and machine learning models to forecast water quality metrics \cite{maier}; but, in recent years, researchers have shown that machine learning models are particularly well suited to take advantage of recent advancements in computer science and the rising availability of water quality data  \cite{Hanson, zwart}. A shortcoming of past limnological forecasting studies that support the use of machine learning is that they tend to focus on a selection of sites that are not representative of freshwater systems across a broad geographic scale. This paper has the primary aim of making a comprehensive comparison of state-of-the-art machine learning methods by evaluating forecasts at 34 different sites across North America at different times of the year.

The time series that are used in this study are taken from the National Ecological Observatory Network's (NEON) Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to data that is collected and made publicly accessible by NEON \cite{Quinn}. A common finding across the challenge is that a day of year historical mean model (also referred to as the climatology model) commonly produces top scoring forecasts \cite{Dietze, Quinn}. For instance, in a model comparison that examined the forecasts for phenological time series, \cite{Dietze} found that the climatology model outperformed all except one of the submitted models; and, the best performing model only marginally outperformed the climatology model. Thus, a primary consideration in this paper is comparing the performance of the machine learning models against the historical null model.

There are many promising neural network architectures that have not been evaluated for water quality forecasting. Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, researchers have identified that LSTM's have notable shortcomings, and there have been continued efforts towards the development of better performing models. Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures. We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently published neural network models.

@@

We evaluated the forecasting performance of 11 different models on water temperature, dissolved oxygen and chlorophyll-A time series, which were collected by in-situ sensors at 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON. There is variability across sites in which target variables were observed. Additionally, maintenance issues led to gaps in the data which also varied with the site. Since the training data spanned less than 3 years, we excluded time series that had gaps longer than a year. Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.

The imputation of missing data proved to be critical to the performance of the ML models. After experimenting with data filling methods that resulted in poor model performance, we developed an imputation method inspired by the climatology model. If the gap size was less than 5 days, then the gap was filled using a Gaussian Process Filter. For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.

We compared the performance of the neural network models to the climatology and naive persistence model. The climatology model generates forecasts by finding the daily mean and standard deviation and draws samples from a Gaussian Distribution with these parameters. The naive persistence model finds the last observed value of the target variable and predicts this value for each day in the forecast horizon. For each model, we computed the Continuous Ranked Probability Score (CRPS) and the Root Mean Square Error (RMSE). To gauge how models performed in relation to the climatology model, we computed the Continuous Ranked Probability Skill Score (CRPSS) which we defined as 
\begin{equation}
CRPSS_{model} = 1 - \frac{CRPS_{model}}{CRPS_{clim.}}.
\end{equation}

And using the naive persistence model as a reference, we computed the RMSE Skill Score,

\begin{equation}
RMSE-SS_{model} = 1 - \frac{RMSE_{model}}{RMSE_{naive.}}.
\end{equation}

If a target model outperforms than the reference model, then the target model will have a positive skill score. If the target model performs worse than the reference model, the skill score will be negative.

Talk about theta model

For the 8 neural network models that were investigated, there were varying design choices made regarding which type of covariates could be used. For instance, models could accept past covariates only, future covariates only or both future and past covariates.

For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates. For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate. The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

@@

Models tend to struggle to accurately forecast chlorophyll during bloom conditions \cite{woelmer}. Accurately predicting timing and magnitude of bloom events is not surprising
